On the Synchronization Power of the Openflow Data Plane

### 2015-04-19 inband-sync.tex
final for now

\newcommand{\FlowMod}{\emph{FlowMod}\xspace}
\newcommand{\match}{\emph{match}\xspace}
\newcommand{\action}{\emph{action}\xspace}
\newcommand{\flags}{\emph{flags}\xspace}
\newcommand{\checko}{\texttt{chkov}\xspace}

@@

\begin{abstract}
Control planes of modern Software-Defined Networks (SDNs)
are \emph{distributed systems}: to ensure availability and fault-tolerance,
to improve load-balancing, and to reduce overheads,
control modules are physically distributed.
%a redundant control plane
%can ensure a high availability, and a spatially distributed
%control plane allows to handle time-critical dataplane
%events close to their origins.
%[[PK this is a contraversial point, not important for us
%Moreover, an SDN may be managed by multiple administrators simultaneously.
%]]
However, a distributed control plane also introduces
new challenges, and in order to guarantee a consistent network operation,
the actions of different controllers may need to be synchronized.
%[[PK sounds contradicting to the fact that such systems already exist
%However, not much is known today about how to implement
%such a distributed control plane.
%]]
%
A well-known concept to solve synchronization and consistency problems
in distributed systems are \emph{transactions}: a transaction defines a sequence
of operations which are performed in an atomic, serializable manner, providing all-or-nothing semantics.
This paper shows that powerful transactions can also be implemented in software-defined networks,
using the standard Openflow protocol.
%Indeed, recent Openflow versions already come with useful but primitive
%features, in particular \emph{bundling} and \emph{rule-overlap checks}.
In particular, we present a
transactional interface which supports arbitrary atomic
transactions
over the switch configuration space: Our transactions
can include complex notions of conflicts (beyond flow overlap checks),
as well as computations and multiple controller interactions. 
We show how our interface can be used for applications such as 
in-band controller consensus, consistent policy installation, or load-balancing.
\end{abstract}

@@

In order to avoid conflicts, \emph{FlowMod} can be used together with a the $\checko$ flag ($CHECK\_OVERLAP$):
if the flow spaces of an old
and the new rule overlap,
that is, if there is an entry with an overlapping match part but a
different action,
 and if the two rules have the same priority, then
\emph{FlowMod} will fail.

@@

In this paper, we will use the notation
%$\textbf{bundle}\{\;\ldots\;\}\rightarrow res$ to refer to represent a
$\textbf{bundle}\{\;\ldots\;\}$ to refer to a
sequence of commands sent in a bundle. The bundle returns 
$\top$ if and only if 
all commands are performed successfully, and otherwise $\bot$.
The bundles used in this paper will always have the following structure:
\begin{enumerate}
\item First, it contains a \emph{create bundle} message: a \textsf{OFPT\_BUNDLE\_CONTROL} message of type
\textsf{OFPBCT\_OPEN\_REQUEST}.
\item The follows a sequence of commands (e.g., \emph{FlowMod}), each wrapped in an 
\emph{add to bundle message} (\textsf{OFPT\_BUNDLE\_\add\_MESSAGE}) with the \emph{bundle identifier}.
\item Finally, a \emph{commit bundle} control message (\textsf{OFPT\_BUNDLE\_CONTROL}) of 
type \textsf{OFPBCT\_COMMIT\_REQUEST}.
\end{enumerate}
Again, the bundle always returns a binary result: \emph{error} ($\bot$) or \emph{success} ($\top$).

@@

\hide{
To commit the transaction,  the controller sends a \textsf{OFPT\_BUNDLE\_CONTROL} message with type
\textsf{OFPBCT\_COMMIT\_REQUEST} and waits until a matching control
message from the switch is received. When the switch receives the
commit request it tries to atomically process all the buffered messages
equipped with the corresponding bundle identifier and, if the
processing is successful for all of them, returns  a \textsf{OFPT\_BUNDLE\_CONTROL} message with type
\textsf{OFPBCT\_CLOSE\_REPLY}; otherwise it sends a specific error
message  pointing out the aborted transactional operation.

Algorithm~\ref{alg:commit} describes how to commit
the bundle.
\begin{algorithm}[h]
    \caption{$\textit{try-commit}()$}
    \label{alg:commit}
    \begin{algorithmic}[1]
    \Require  current bundle id $bid$.
		    \State $cmd \gets struct ofp_bundle_ctrl_msg$
    		\State  $cmd.bundle\_id \gets bid$
    		\State  $cmd.type \gets $ \textsf{OFPBCT\_COMMIT\_REQUEST}
    		\State  $cmd.flags \gets $ \textsf{OFPBF\_ATOMIC}
    		\State send $cmd$
		\Return $cmd$
    \end{algorithmic}
\end{algorithm}
}

@@

At first sight, the Openflow bundle feature seems to provide everything
we need to implement our transactional interface: the bundle
 defines a sequence of read and write operations
 on the switch state which are executed atomically.
 However, while the bundle can indeed be seen 
 as a ``mini-transaction'', it comes with several
 limitations, which we discuss in the following.

First, if employed without any additional features, a bundle does not allow
 to change the switch state in any non-trivial way:
the read and write actions in the bundle do
 not support any conditional modifications or computations which
 depend on the read results. 
But also using the bundle together with the check overlap feature ($\checko$)
comes with limitations: while it is possible
 to \emph{abort} transactions depending on certain conditions
 in the \emph{overlapping
 flow spaces}, oftentimes it is desirable to be able
 to define more general conditions and notions of conflicts
 under which transactions should not take effect. Moreover,
 sometimes it is desirable to react to conflicts in smarter ways than
 to simply abort, e.g.,
 by supporting \emph{modifications}.
Let us consider two examples.

\begin{enumerate}
\item \textbf{Example 1: More complex dependencies.} Let us consider a controller application in charge
of \emph{load-balancing}. 
In a load-balancing application, seemingly independent actions, defined over
completely independent logical flow spaces,
may in fact be dependent due to collocation in the underlying physical network.
To be more concrete, say that the objective of the load-balancer
is to distribute flows equally across links or servers.
If multiple controllers concurrently and independently
update forwarding rules (say, two ``independent'' flows with source IP address ranges $1*$ and $0*$), 
they may unvoluntarily
unbalance the flow allocation, as the managed flows are embedded
on the same physical links.  
Flow space overlap checks cannot be used to detect
such conflicts.

@@

\item \textbf{Example 2: Supporting modifications.} Consider the problem of modifying previously
installed forwarding policies.
While a bundle allows to conditionally add new rules or remove
old rules (and, using the flow space overlap
check feature $\checko$, to abort the transaction if potentially conflicting
rules were already there), it cannot be used
to modify rules in non-trivial ways.
In other words, the bundle and overlap check only helps
for the conditional creation of rules, not their modification.
\end{enumerate}

@@

Our approach is based on the distributed management of \emph{identifiers (IDs)}. In particular, 
we introduce two new synchronization objects (and describe their implementation in standard Openflow),
which allows controllers to coordinate 
themselves based on these identifiers.

@@

With these base objects, we provide the following shared memory abstraction, with 
write-compare access semantics:
\begin{itemize}
\item By performing $\textit{write}(\addr,k)$,
  $\addr,k\in\Nat$, the value at memory offset $\addr$ is set to $k$. For
  example, the offset $\addr$ may be used to identify the policy (identifier),
  but our approach is more general.

\item The operation $\textit{\compare}(\addr,k)$, $\addr,k\in\Nat$,
aborts current transaction, if the value at memory offset $\addr$ is not equal to $k$.
%returns $\textit{true}$ if and
%  only if $k$ is not currently claimed by any controller; if
%  $\textit{false}$ is returned, we say that the check operation
%  \emph{fails} (which causes an abort of the transaction invoking it).
\end{itemize}

As we will see, the bundle feature is not required to implement this abstraction.
Also note that memory read operations correspond to reading the switch 
configuration, and are simple as they do not affect the execution of the transaction 
(unlike the $\textit{\compare}$ operation).

With this memory interface we can immediately implement 
a powerful \emph{compare-and-set} (CAS) operation which takes two integer parameters
$\textit{old}$ and $\textit{new}$, reads the register and, if the
value of it is $\textit{old}$, replaces it with $\textit{new}$ and
returns $\top$; otherwise, $\bot$ is returned and
we say that the CAS operation \emph{fails} (which causes an abort if
the transaction invoking it).

@@

%The postulated container object can store an arbitrary number of \emph{items}.
%As we will see, these items correspond to flow entries on the Openflow switch.
%The same item may even exist multiple times in the container (i.e., the container
%is a multi-set).
%The container
%provides two main functions:
\begin{itemize}
\item With $\textit{claim}(k)$, $i\in[N]$,
  $k\in\Nat$, a controller $i$ \emph{claims}
  identifier $k$.

\item With $\textit{unclaim}(k)$, $i\in[N]$, $k\in\Nat$, controller $i$
  \emph{unclaims} identifier $k$;

@@

In order to check if a number is claimed, we try to create the 
corresponding entry while setting the
flag \textsf{\checko}, and using different actions ($2$ instead of $1$), thereby inflicting 
a failure in case an entry with overlapping match value exists. 
If the check succeeds, we delete the entry. Considering $m_1$ and $m_2$ as defined above, 
and their respective actions $a_1=x_1\concat c1$ and $a_2=x_2\concat c2$,  they overlap 
with different actions iff $x_1=x_2$. Therefore, if one is used in \claimcheck after the other used in the claim operation, 
a conflict is detected iff $x_1=x_2$, as desired.

Note that for switch space reasons, when checking for claims, also the deletion of the added tester flow entry is included,
which however has no effect on claims and checks of other controllers; hence, the deletion command does not 
have to be sent as part of a bundle. Therefore, the \textsf{\claimcheck} operation is atomic by nature, 
similar to the \textsf{claim} and \textsf{unclaim} operations, and is implemented as essentially one command.

@@

\begin{algorithm}[H]
    \caption{$\textit{\claimcheck}(x)$}
    \label{alg:check}
    \begin{algorithmic}[1]
    \Require \emph{self} as the calling controller ID.
%    \Ensure
    		\State $value \gets x\concat x$
    		\State $mask \gets self\concat all-1$
    		\State $match \gets (value,mask)$
    		%\State $cookie \gets x\concat self$
    		\State $action \gets 2$
    		\State $flag \gets \textsf{OFPFF\_\checko}$
    		\State $cmd1\gets \emph{FlowMod}(match, op = \add, flag, action) $
    		\State $cmd2\gets \emph{FlowMod}(match, op = \dele) $
			\Return $cmd1$,$cmd2$
    \end{algorithmic}
\end{algorithm}

@@

In order to check if a value is written in a given address, 
we try to add it the same way as in the \memwrite operation,  
but we also set the \textsf{\checko} flag. If the value is not there, 
our action differs from the action of the existing entry, 
thereby inflicting a failure.
In case the value is there, our flow entry replaces the existing flow entry 
with an identical one.
In addition, we present in Algorithm \ref{alg:cas} a simple implementation 
of \emph{compare-and-set (CAS)}, using the two last operations.

@@

\begin{algorithm}[h]
    \caption{$\textit{\compare}(\addr,k)$}
    \label{alg:compare}
    \begin{algorithmic}[1]
    %\Require  id $i$.
%    \Ensure
    		\State $value \gets \addr$
    		\State $mask \gets  all-1$
    		\State $match \gets (value,mask)$
    		%\State $cookie \gets x\concat self$
    		\State $action \gets k$
    		\State $flag \gets \textsf{OFPFF\_\checko}$
    		\State $cmd\gets \emph{FlowMod}(match, op = \add, flag, action) $
			\Return $cmd$
    \end{algorithmic}
\end{algorithm}

@@

As a second example, let us consider a simple policy composition protocol that
allows a set of  controllers to concurrently update switch  \emph{policies} , i.e., sets of
effective flow tables rules, under the
condition that these rules can be composed with the currently installed
policy. 
Let us assume that the composition predicate is encoded in a user-specific
$\textit{apply\_update}()$ function~\cite{cpc}.

@@

We use increasing numbers as policies
ids and use them in order to make sure that the new policy, with id
$x$, replaces the (what considered as) current policy, with id
$x-1$. Considering the policies ids, we send the policy installation
commands as part of a bundle that also tries to perform
$CAS(x-1,x)$, thereby insuring that the commands are
performed iff the current policy is still there and preventing other
controllers to replace the new policy with another without observing it.

@@

Algorithm~\ref{alg:update} claims the current policy id in order to prevent someone of using it with different policy in the future (during the current update) and in the bundle used to update the policy it also checks that the new policy id that it about to use (chosen from a finite set of ids) wasn't claimed by someone else.

@@

\begin{algorithm}[t]
    \caption{Policy update with only CAS}
    \label{alg:simple-update}
    \begin{algorithmic}[1]
    \Require unsafe policy update function $U$, a predefined switch (memory) address for policy id: $\pid_{\addr}$
    \Ensure installed policy is consistent with previous one
 		%\Do
 		\Repeat
 			\State $\pid,policy\gets$ \textbf{read} switch configuration
 			\State $cmds\gets U(policy)$
 			\startTxn
	 			\State $CAS(\pid_{\addr},\pid,\pid+1)$
	 			\State \textbf{send} $cmds$ %conditioned on $metadata=my\_a$
	 			%\State update policy selector to $my\_a$
 			\endTxn
     		%\State $res \gets $ commit transaction
     	\Until{$res\neq\texttt{abort}$}
     	%\doWhile{$res=False$}
%   \vspace{3mm}
			\Return

@@

\begin{algorithm}[t]
    \caption{Advanced policy update}
    \label{alg:update}
    \begin{algorithmic}[1]
        \Require unsafe policy update function $U$, a predefined switch (memory) address for policy id: $\pid_{\addr}$, a predefined id space $C$.
    \Ensure installed policy is consistent with previous one
 		%\Do
 		\Repeat
		 	\State $\pid,claims,policy\gets$ \textbf{read} switch configuration
 			\State $claim(\pid)$
 			\State $\pid 2,claims,policy\gets$ \textbf{read} switch configuration
 			\If {$\pid\neq \pid 2$}
	 			\State $unclaim(\pid)$
 				\State {\bf continue} (\textbf{restart} loop)
 			\EndIf
 			\State $my\_id\gets$ choose a number from $C\setminus claims$
 			\State $cmds\gets U(policy)$
 			\startTxn
 				\State $\claimcheck(my\_id)$
	 			\State $CAS(\pid_{\addr}, \pid,my\_id)$
	 			\State \textbf{send} $cmds$ %conditioned on $metadata=my\_a$
	 			%\State update policy selector to $my\_a$
 			\endTxn
	 		\State $unclaim(\pid)$
     		%\State $res \gets $ commit transaction
     	\Until{$res\neq\texttt{abort}$}
     	%\doWhile{$res=False$}
%   \vspace{3mm}
			\Return

@@

\lstset{% general command to set parameter(s)
	basicstyle=\small, % print whole listing small
	keywordstyle=\color{black}\bfseries, % underlined bold black keywords
	identifierstyle=, % nothing happens
	commentstyle=\color{mygrey}, % white comments
	stringstyle=\ttfamily, % typewriter type for strings
	showstringspaces=false, % no special string spaces
	language=C}
\begin{lstlisting}
struct ofp_flow_mod {
	struct ofp_header header;
	uint64_t cookie; /* Opaque controller-issued identifier. */
	uint64_t cookie_mask; /* Mask used to restrict the cookie bits
		that must match when the command is
		OFPFC_MODIFY* or OFPFC_\dele*. A value
		of 0 indicates no restriction. */
	/* Flow actions. */
	uint8_t table_id; /* ID of the table to put the flow in.
		For OFPFC_\dele_* commands, OFPTT_ALL
		can also be used to delete matching
		flows from all tables. */
	uint8_t command; /* One of OFPFC_*. */
	uint16_t idle_timeout; /* Idle time before discarding (seconds). */
	uint16_t hard_timeout; /* Max time before discarding (seconds). */
	uint16_t priority; /* Priority level of flow entry. */
	uint32_t buffer_id; /* Buffered packet to apply to, or
		OFP_NO_BUFFER.
		Not meaningful for OFPFC_\dele*. */
	uint32_t out_port; /* For OFPFC_\dele* commands, require
		matching entries to include this as an
		output port. A value of OFPP_ANY
		indicates no restriction. */
	uint32_t out_group; /* For OFPFC_\dele* commands, require
		matching entries to include this as an
		output group. A value of OFPG_ANY
		indicates no restriction. */
	uint16_t flags; /* Bitmap of OFPFF_* flags. */
	uint16_t importance; /* Eviction precedence (optional). */
	struct ofp_match match; /* Fields to match. Variable size. */
	/* The variable size and padded match is always followed by instructions. */
	//struct ofp_instruction_header instructions[0];
		/* Instruction set - 0 or more. The length
		of the instruction set is inferred from
		the length field in the header. */
};
\end{lstlisting}

@@

%/* Message structure for OFPT_BUNDLE_CONTROL. */
\begin{lstlisting}
struct ofp_bundle_ctrl_msg {
	struct ofp_header header;
	uint32_t bundle_id; /* Identify the bundle. */
	uint16_t type; /* OFPBCT_*. */
	uint16_t flags; /* Bitmap of OFPBF_* flags. */
	/* Bundle Property list. */
	struct ofp_bundle_prop_header properties[0]; /* Zero or more properties. */
};
\end{lstlisting}

@@

\begin{algorithm}[h]
    \caption{Policy composition without bundle}
    \label{alg:wobundle}
    \begin{algorithmic}[1]
    \Require policy $P$ (a set of FlowMod entry creation commands), three consecutive flow tables ids (per controller) $T_s,T_0,T_1$, active table index $active\in\{0,1\}$.
    %\Ensure installed policy is consistent with previous one
    \State $unactive \gets active + 1 \mod{2}$
    \For {$cmd \in P$}
	    \State $cmd.action\gets Reencode(cmd.action)$
	    \State $cmd.table\_id\gets T_{unactive}$
	    \State send $cmd$
    \EndFor
    \State modify the action of $T_s$ entry to goto $T_{unactive}$
    \State clear table $T_{active}$
    \State $active \gets unactive$
	\Return

@@

    		\State $op \gets \textsf{OFPGC\_\dele}$
    		\State $group\_id \gets old$
    		\State $cmd1\gets GroupMod(op, group\_id) $

>>>>

\newcommand{\FlowMod}{\emph{FlowMod}\xspace}
\newcommand{\match}{\emph{match}\xspace}
\newcommand{\action}{\emph{action}\xspace}
\newcommand{\flags}{\emph{flags}\xspace}
\newcommand{\checko}{\texttt{check\_overlap}\xspace}

@@

\begin{abstract}
Control planes of modern Software-Defined Networks (SDNs)
are \emph{distributed systems}: to ensure availability and fault-tolerance,
to improve load-balancing, and to reduce overheads,
control modules are physically distributed.
%a redundant control plane
%can ensure a high availability, and a spatially distributed
%control plane allows to handle time-critical dataplane
%events close to their origins.
%[[PK this is a contraversial point, not important for us
%Moreover, an SDN may be managed by multiple administrators simultaneously.
%]]
However, a distributed control plane also introduces
new challenges, and in order to guarantee a consistent network operation,
the actions of different controllers may need to be synchronized.
%[[PK sounds contradicting to the fact that such systems already exist
%However, not much is known today about how to implement
%such a distributed control plane.
%]]
%
A well-known concept to solve synchronization and consistency problems
in distributed systems are \emph{transactions}: a transaction defines a sequence
of operations which are performed in an atomic, serializable manner, providing all-or-nothing semantics.
This paper shows that powerful transactions can also be implemented in software-defined networks,
using the standard Openflow protocol.
%Indeed, recent Openflow versions already come with useful but primitive
%features, in particular \emph{bundling} and \emph{rule-overlap checks}.
In particular, we present a
transactional interface which supports arbitrary atomic
transactions
over the switch configuration space: Our transactions
can include complex notions of conflicts (beyond flow overlap checks),
as well as computations and multiple controller interactions. 
We discuss applications of our transactional memory interface to  
in-band controller consensus, consistent policy installation, and load-balancing.
\end{abstract}

@@

In order to avoid conflicts, \emph{FlowMod} can be used together with a the $\checko$ flag:
if the flow spaces of an old
and the new rule overlap,
that is, if there is an entry with an overlapping match part but a
different action,
 and if the two rules have the same priority, then
\emph{FlowMod} will fail.

@@

In this paper, we will use the notation
%$\textbf{bundle}\{\;\ldots\;\}\rightarrow res$ to refer to represent a
$\textbf{bundle}\{\;\ldots\;\}$ to refer to a
sequence of commands sent in a bundle. The bundle returns 
$\top$ if and only if 
all commands are performed successfully, and otherwise $\bot$.
The bundles used in this paper will always have the following structure:
\begin{enumerate}
\item First, it contains a \emph{create bundle} message: a \texttt{ofpt\_bundle\_control} message of type
\texttt{OFPBCT\_OPEN\_REQUEST}.
\item The follows a sequence of commands (e.g., \emph{FlowMod}), each wrapped in an 
\emph{add to bundle message} (\texttt{ofpt\_bundle\_add\_message}) with the \emph{bundle identifier}.
\item Finally, a \emph{commit bundle} control message (\texttt{ofpt\_bundle\_control}) of
type \texttt{ofpbct\_commit\_request}.
\end{enumerate}
Again, the bundle always returns a binary result: \emph{error} ($\bot$) or \emph{success} ($\top$).

@@

\hide{
To commit the transaction,  the controller sends a \texttt{OFPT\_BUNDLE\_CONTROL} message with type
\texttt{OFPBCT\_COMMIT\_REQUEST} and waits until a matching control
message from the switch is received. When the switch receives the
commit request it tries to atomically process all the buffered messages
equipped with the corresponding bundle identifier and, if the
processing is successful for all of them, returns  a \texttt{OFPT\_BUNDLE\_CONTROL} message with type
\texttt{OFPBCT\_CLOSE\_REPLY}; otherwise it sends a specific error
message  pointing out the aborted transactional operation.

Algorithm~\ref{alg:commit} describes how to commit
the bundle.
\begin{algorithm}[h]
    \caption{$\textit{try-commit}()$}
    \label{alg:commit}
    \begin{algorithmic}[1]
    \Require  current bundle id $bid$.
		    \State $cmd \gets struct ofp_bundle_ctrl_msg$
    		\State  $cmd.bundle\_id \gets bid$
    		\State  $cmd.type \gets $ \texttt{ofpbct\_commit\_request}
    		\State  $cmd.flags \gets $ \texttt{ofpbf\_atomic}
    		\State \textbf{send} $cmd$
		\Return $cmd$
    \end{algorithmic}
\end{algorithm}
}

@@

At first sight, the Openflow bundle feature seems to provide everything
we need to implement our transactional interface: the bundle
 defines a sequence of read and write operations
 on the switch state which is executed atomically.
 However, while the bundle can indeed be seen 
 as a ``mini-transaction'', it comes with several
 limitations, which we discuss in the following.

First, if employed without any additional features, a bundle does not allow
 to change the switch state in any non-trivial way:
the read and write actions in the bundle do
 not support any conditional \emph{modifications} or computations which
 depend on the read results. 
But also using the bundle together with the check overlap feature ($\checko$)
comes with limitations: while it is possible
 to \emph{abort} transactions depending on certain conditions
 in the \emph{overlapping
 flow spaces}, oftentimes it is desirable to be able
 to define more general conditions and notions of conflicts
 under which transactions should not take effect. Moreover,
 sometimes it is desirable to react to conflicts in smarter ways than
 to simply abort, e.g.,
 by supporting \emph{modifications}.
Let us consider two examples.

\begin{enumerate}
\item \textbf{Example 1: More complex dependencies.} Let us consider two controllers 
in charge
of \emph{load-balancing}. 
In a load-balancing application, seemingly independent actions, defined over
completely independent logical flow spaces (say, two flows with source IP address ranges $1*$ and $0*$),
may in fact be dependent due to collocation in the underlying physical network.
If the two controllers concurrently and independently
update forwarding rules,  
they may unvoluntarily
unbalance the flow allocation, as the managed flows are embedded
on the same physical links.  
Flow space overlap checks cannot be used to detect
such conflicts.

@@

\item \textbf{Example 2: Supporting modifications.} Consider the problem of modifying previously
installed forwarding policies.
While a bundle allows to conditionally add new rules or remove
old rules (and, using the flow space overlap
check feature $\checko$, to abort the transaction if potentially conflicting
rules were already there), it cannot be used
to modify rules in non-trivial ways.
In other words, the bundle and overlap check only help
for the \emph{conditional creation} of rules, not their \emph{modification}.
\end{enumerate}

@@

Our approach is based on the distributed management of \emph{identifiers (IDs)}. In particular, 
we introduce two new synchronization objects (and describe their implementation in standard Openflow),
which allow controllers to coordinate 
themselves based on these identifiers.

@@

With these base objects, we provide the following shared memory abstraction, with 
write-compare access semantics:
\begin{itemize}
\item By performing $\textit{write}(\addr,k)$,
  $\addr,k\in\Nat$, the value at memory offset $\addr$ is set to $k$. For
  example, the offset $\addr$ may correspond to the policy identifier (\emph{pid}),
  but our approach is more general.

\item By performing an operation $\textit{\compare}(\addr,k)$, $\addr,k\in\Nat$,
a transaction can be aborted if the value at memory offset $\addr$ is not equal to $k$.
%returns $\textit{true}$ if and
%  only if $k$ is not currently claimed by any controller; if
%  $\textit{false}$ is returned, we say that the check operation
%  \emph{fails} (which causes an abort of the transaction invoking it).
\end{itemize}

As we will see, the bundle feature is not required to implement these semantics.
Also note that memory read operations correspond to reading the switch 
configuration, and are simple as they do not affect the execution of the transaction 
(unlike the $\textit{\compare}$ operation).

Given this memory interface, we can immediately implement 
a powerful \emph{Compare-And-Set} (CAS) operation which takes two integer parameters
$\textit{old}$ and $\textit{new}$, reads the register and, if the
value of it is $\textit{old}$, replaces it with $\textit{new}$ and
returns $\top$; otherwise, $\bot$ is returned and
we say that the CAS operation \emph{fails} (which causes an abort of
the invoking transaction).

@@

%The postulated container object can store an arbitrary number of \emph{items}.
%As we will see, these items correspond to flow entries on the Openflow switch.
%The same item may even exist multiple times in the container (i.e., the container
%is a multi-set).
%The container
%provides two main functions:
\begin{itemize}
\item With $\textit{claim}(k)$, 
  $k\in\Nat$, a controller $i\in[N]$ \emph{claims}
  identifier $k$.

\item With $\textit{unclaim}(k)$, $k\in\Nat$, controller $i\in[N]$
  \emph{unclaims} identifier $k$.

@@

In order to check if a number is claimed, we try to create the 
corresponding entry while setting the
flag \texttt{\checko}, and using different actions ($2$ instead of $1$), thereby inflicting
a failure in case an entry with overlapping match value exists. 
If the check succeeds, we delete the entry. Considering $m_1$ and $m_2$ as defined above, 
and their respective actions $a_1=x_1\concat c1$ and $a_2=x_2\concat c2$,  they overlap 
with different actions iff $x_1=x_2$. Therefore, if one is used in \claimcheck after the other used in the claim operation, 
a conflict is detected iff $x_1=x_2$, as desired.

Note that for switch space reasons, when checking for claims, also the deletion of the added tester flow entry is included,
which however has no effect on claims and checks of other controllers; hence, the deletion command does not 
have to be sent as part of a bundle. Therefore, the \texttt{\claimcheck} operation is atomic by nature,
similar to the \texttt{claim} and \texttt{unclaim} operations, and is implemented as essentially one command.

@@

\begin{algorithm}[H]
    \caption{$\textit{\claimcheck}(x)$}
    \label{alg:check}
    \begin{algorithmic}[1]
    \Require \emph{self} as the calling controller ID.
%    \Ensure
    		\State $value \gets x\concat x$
    		\State $mask \gets self\concat all-1$
    		\State $match \gets (value,mask)$
    		%\State $cookie \gets x\concat self$
    		\State $action \gets 2$
    		\State $flag \gets \texttt{OFPFF\_\checko}$
    		\State $cmd1\gets \emph{FlowMod}(match, op = \add, flag, action) $
    		\State $cmd2\gets \emph{FlowMod}(match, op = \dele) $
			\Return $cmd1$,$cmd2$
    \end{algorithmic}
\end{algorithm}

@@

In order to check if a value is written in a given address, 
we try to add it the same way as in the \memwrite operation,  
but we also set the \texttt{\checko} flag. If the value is not there,
our action differs from the action of the existing entry, 
thereby inflicting a failure.
In case the value is there, our flow entry replaces the existing flow entry 
with an identical one.
In addition, we present in Algorithm \ref{alg:cas} a simple implementation 
of \emph{Compare-And-Set (CAS)}, using the two last operations.

@@

\begin{algorithm}[h]
    \caption{$\textit{\compare}(\addr,k)$}
    \label{alg:compare}
    \begin{algorithmic}[1]
    %\Require  id $i$.
%    \Ensure
    		\State $value \gets \addr$
    		\State $mask \gets  all-1$
    		\State $match \gets (value,mask)$
    		%\State $cookie \gets x\concat self$
    		\State $action \gets k$
    		\State $flag \gets \texttt{OFPFF\_\checko}$
    		\State $cmd\gets \emph{FlowMod}(match, op = \add, flag, action) $
			\Return $cmd$
    \end{algorithmic}
\end{algorithm}

@@

As a second example, let us consider a simple policy composition protocol that
allows a set of  controllers to concurrently update switch  \emph{policies}, i.e., sets of
effective flow tables rules, under the
condition that these rules can be composed with the currently installed
policy. 
Let us assume that the composition predicate is encoded in a user-specific
$\textit{apply\_update}()$ function~\cite{cpc}.

@@

We use increasing numbers as policies
IDs and use them in order to make sure that the new policy, with id
$x$, replaces the (what considered as) current policy, with id
$x-1$. Considering the policies IDs, we send the policy installation
commands as part of a bundle that also tries to perform
$CAS(x-1,x)$, thereby insuring that the commands are
performed iff the current policy is still there and preventing other
controllers to replace the new policy with another without observing it.

@@

Algorithm~\ref{alg:update} claims the current policy id in order to prevent someone of using it with different policy in the future (during the current update) and in the bundle used to update the policy it also checks that the new policy id that it about to use (chosen from a finite set of IDs) wasn't claimed by someone else.

@@

\begin{algorithm}[t]
    \caption{Policy update with only CAS}
    \label{alg:simple-update}
    \begin{algorithmic}[1]
    \Require unsafe policy update function $U$, a predefined switch (memory) address for policy id: $\pid_{\addr}$
    \Ensure installed policy is consistent with previous one
 		%\Do
 		\Repeat
 			\State $\pid,policy\gets$ \textbf{read} switch configuration
 			\State $cmds\gets U(policy)$
 			\startTxn
	 			\State $CAS(\pid_{\addr},\pid,\pid+1)$
	 			\State \textbf{send} $cmds$ %conditioned on $metadata=my\_a$
	 			%\State update policy selector to $my\_a$
 			\endTxn
     		%\State $res \gets $ commit transaction
     	\Until{$res\neq \bot$}
     	%\doWhile{$res=False$}
%   \vspace{3mm}
			\Return

@@

\begin{algorithm}[t]
    \caption{Advanced policy update}
    \label{alg:update}
    \begin{algorithmic}[1]
        \Require unsafe policy update function $U$, a predefined switch (memory) address for policy id: $\pid_{\addr}$, a predefined id space $C$.
    \Ensure installed policy is consistent with previous one
 		%\Do
 		\Repeat
		 	\State $\pid,claims,policy\gets$ \textbf{read} switch configuration
 			\State $claim(\pid)$
 			\State $\pid 2,claims,policy\gets$ \textbf{read} switch configuration
 			\If {$\pid\neq \pid 2$}
	 			\State $unclaim(\pid)$
 				\State {\bf continue} (\textbf{restart} loop)
 			\EndIf
 			\State $my\_id\gets$ choose a number from $C\setminus claims$
 			\State $cmds\gets U(policy)$
 			\startTxn
 				\State $\claimcheck(my\_id)$
	 			\State $CAS(\pid_{\addr}, \pid,my\_id)$
	 			\State \textbf{send} $cmds$ %conditioned on $metadata=my\_a$
	 			%\State update policy selector to $my\_a$
 			\endTxn
	 		\State $unclaim(\pid)$
     		%\State $res \gets $ commit transaction
     	\Until{$res\neq\bot$}
     	%\doWhile{$res=False$}
%   \vspace{3mm}
			\Return

@@

\lstset{% general command to set parameter(s)
	basicstyle=\small, % print whole listing small
	keywordstyle=\color{black}\bfseries, % underlined bold black keywords
	identifierstyle=, % nothing happens
	commentstyle=\color{mygrey}, % white comments
	stringstyle=\ttfamily, % typewriter type for strings
	showstringspaces=false, % no special string spaces
	language=C}
\begin{lstlisting}
struct ofp_flow_mod {
	struct ofp_header header;
	uint64_t cookie; /* Opaque controller-issued identifier. */
	uint64_t cookie_mask; /* Mask used to restrict the cookie bits
		that must match when the command is
		\texttt{ofpfc_{modify}*} or \texttt{ofpfc_{delete}*. A value
		of 0 indicates no restriction. */
	/* Flow actions. */
	uint8_t table_id; /* ID of the table to put the flow in.
		For OFPFC_DELETE_* commands, OFPTT_ALL
		can also be used to delete matching
		flows from all tables. */
	uint8_t command; /* One of OFPFC_*. */
	uint16_t idle_timeout; /* Idle time before discarding (seconds). */
	uint16_t hard_timeout; /* Max time before discarding (seconds). */
	uint16_t priority; /* Priority level of flow entry. */
	uint32_t buffer_id; /* Buffered packet to apply to, or
		OFP_NO_BUFFER.
		Not meaningful for OFPFC_\dele*. */
	uint32_t out_port; /* For OFPFC_\dele* commands, require
		matching entries to include this as an
		output port. A value of OFPP_ANY
		indicates no restriction. */
	uint32_t out_group; /* For OFPFC_\dele* commands, require
		matching entries to include this as an
		output group. A value of OFPG_ANY
		indicates no restriction. */
	uint16_t flags; /* Bitmap of OFPFF_* flags. */
	uint16_t importance; /* Eviction precedence (optional). */
	struct ofp_match match; /* Fields to match. Variable size. */
	/* The variable size and padded match is always followed by instructions. */
	//struct ofp_instruction_header instructions[0];
		/* Instruction set - 0 or more. The length
		of the instruction set is inferred from
		the length field in the header. */
};
\end{lstlisting}

@@

%/* Message structure for OFPT_BUNDLE_CTRL. */
\begin{lstlisting}
struct ofp_bundle_ctrl_msg {
	struct ofp_header header;
	uint32_t bundle_id; /* Identify the bundle. */
	uint16_t type; /* OFPBCT_*. */
	uint16_t flags; /* Bitmap of OFPBF_* flags. */
	/* Bundle Property list. */
	struct ofp_bundle_prop_header properties[0]; /* Zero or more properties. */
};
\end{lstlisting}

@@

\begin{algorithm}[h]
    \caption{Policy composition without bundle}
    \label{alg:wobundle}
    \begin{algorithmic}[1]
    \Require policy $P$ (a set of FlowMod entry creation commands), three consecutive flow tables IDs (per controller) $T_s,T_0,T_1$, active table index $active\in\{0,1\}$.
    %\Ensure installed policy is consistent with previous one
    \State $unactive \gets active + 1 \mod{2}$
    \For {$cmd \in P$}
	    \State $cmd.action\gets Reencode(cmd.action)$
	    \State $cmd.table\_id\gets T_{unactive}$
	    \State send $cmd$
    \EndFor
    \State modify the action of $T_s$ entry to goto $T_{unactive}$
    \State clear table $T_{active}$
    \State $active \gets unactive$
	\Return

@@

    		\State $op \gets \texttt{OFPGC\_\dele}$
    		\State $group\_id \gets old$
    		\State $cmd1\gets GroupMod(op, group\_id) $

    		\State $op \gets \texttt{OFPGC\_\add}$
    		\State $type \gets \texttt{OFPGT\_INDIRECT}$
    		\State $group\_id \gets new$
    		\State $buckets \gets [dummy\_backet]$
    		\State $cmd2\gets GroupMod(op, type, group\_id, buckets) $
			\Return $cmd1,cmd2$
    \end{algorithmic}
\end{algorithm}
